Manuscript figures

### 2025-07-07 Manuscript_figures.Rmd
Update for revision

---
title: "Manuscript figures"
author: "Rasmus Rydbirk"
date: "2024-11-11"
output: 
  html_document:
    toc: True
    toc_float: True
---

@@

```{r setup, message=FALSE}
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(pagoda2)
library(scHelper) # github.com/rrydbirk/scHelper
library(cacoa)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ComplexHeatmap)
library(destiny)
library(Seurat)
library(harmony)
library(slingshot)
library(circlize)
library(cowplot)
library(sccore)
library(corrplot)

@@

@@

## Figure 1G

@@

## Figure 1H

@@

## Figure 1I

@@

@@

```{r f2load}
con <- qread("con_vascular.qs", nthreads = 10)
anno.vascular <- qread("anno_vascular.qs")
```

## Figure 2A

@@

## Figure 2B

@@

## Figure 2D

@@

@@

## Figure 3A

@@

## Figure 3B

@@

## Figure 3C

Here, we're plotting proportions of total cells, so we load the Conos object and the annotation for all cells first.
```{r f3c}
con.major <- qread("con_major.qs", nthreads = 10)
anno.major <- qread("anno_major.qs")

@@

## Figure 3D

## Figure 3E

@@

genes <- de[c("ATM","Early LAM","LAM","mono/mac")] %>%
  {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
  lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
  lapply(pull, Gene) %>% 
  {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
  Reduce(c, .) %>% 
  .[match(unique(.), .)]

@@

ha <- ComplexHeatmap::Heatmap(x, 
                              name='Expression', 
                              col=pal, 
                              cluster_columns=FALSE, 
                              show_row_names=FALSE, 
                              show_column_names=FALSE, 
                              top_annotation=tannot,
                              left_annotation=NULL,
                              border=TRUE,
                              show_column_dend = FALSE, 
                              show_row_dend = FALSE,
                              row_km = 5,
                              row_km_repeats = 500,
                              column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")))

@@

## Figures 3F,G

@@

### Figure 3F

@@

### Figure 3G

>>>>

---
title: "Manuscript figures"
author: "Rasmus Rydbirk"
date: "2025-07-07"
output: 
  html_document:
    toc: True
    toc_float: True
---

@@

```{r setup, message=FALSE}
library(conos)
library(qs)
library(dplyr)
library(magrittr)
library(pagoda2)
library(scHelper) # github.com/rrydbirk/scHelper
library(cacoa)
library(ggplot2)
library(rstatix)
library(ggpubr)
library(ComplexHeatmap)
library(destiny)
library(Seurat)
library(harmony)
library(slingshot)
library(circlize)
library(cowplot)
library(sccore)
library(corrplot)
library(RColorBrewer)
library(glmnet)

@@

# Palettes
pal.major <- RColorBrewer::brewer.pal(9, "Set1") %>% 
  setNames(c("Adipocytes", "Myeloid immune cells", "ASPCs", "ECs", "SMCs", "Lymphoid immune cells", "Mast cells", "Lymphatic ECs", "Pericytes")) %>% 
  {setNames(c(., ., sapply(unname(.), \(color) {
    hsv_vals <- rgb2hsv(col2rgb(color))
    new_hues <- hsv(pmin(1, hsv_vals[1] * 0.9), hsv_vals[2], hsv_vals[3])
  })), c(names(.), paste0(names(.), ", female"), paste0(names(.), ", male"))
  )} %>% 
  {setNames(c(., .[c("Lymphoid immune cells", "Myeloid immune cells")]), c(names(.), c("Lymphoid ICs", "Myeloid ICs")))}

@@

## Figures 1b-1f

These figures were made in GraphPad Prism.

## Figure 1g

@@

## Figure 1h

@@

## Figure 1i

@@

## Figure 1k

```{r}
size_data <- read.csv("Adipocyte_size.csv", sep = ",", header = TRUE)
sample_colours <- c("#F8B3BB", "#BD6563", "#7D2A25", "#ABD1D4", "#62649E", "#3E4282")

patient_averages <- size_data %>%
  group_by(ID, Visit, Sex) %>%
  summarise(Average_Diameter = mean(Diameter, na.rm = TRUE)) %>%
  ungroup() %>% 
  mutate(sex_visit = interaction(Visit, Sex) %>% 
           factor(labels = c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")))

plot <- ggplot(patient_averages, aes(x = sex_visit, 
                                     y = Average_Diameter, 
                                     fill = sex_visit)) +
  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
  geom_jitter(size = 1, width = 0.2, alpha = 0.3, show.legend = F) +  # Add individual points
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), 
    text = element_text(family = "serif", size = 12),
    line = element_blank(),
    axis.text.x = element_blank()
  ) +
  scale_fill_manual(values = sample_colours) +
  theme(rect = element_rect(fill = "transparent")) +
  labs(fill = "", x = "", y = "Diameter (µm)")

plot
```

## Figure 1m

```{r}
# Load the CSV file
CLS <- read.delim("Human_WL_CLS_density_for_plotting.csv", sep = ",")

names(CLS)[names(CLS) == "CLS_density_.number_per_mm3."] <- "CLS_density"

# Perform a Kruskal-Wallis test separately for each sex
kw_res_F <- kruskal.test(CLS_density ~ Visit, data = CLS %>% filter(Sex == "F"))
kw_res_M <- kruskal.test(CLS_density ~ Visit, data = CLS %>% filter(Sex == "M"))

# Perform paired Wilcoxon tests for each sex with Holm's correction
paired_wilcox <- function(CLS, sex_group) {
  data_sex <- CLS %>% filter(Sex == sex_group)
  pairwise_tests <- list(
    visit1_vs_visit2 = wilcox.test(
      x = data_sex %>% filter(Visit == "1") %>% pull(CLS_density),
      y = data_sex %>% filter(Visit == "2") %>% pull(CLS_density),
      paired = F
    )$p.value,
    visit1_vs_visit3 = wilcox.test(
      x = data_sex %>% filter(Visit == "1") %>% pull(CLS_density),
      y = data_sex %>% filter(Visit == "3") %>% pull(CLS_density),
      paired = F
    )$p.value,
    visit2_vs_visit3 = wilcox.test(
      x = data_sex %>% filter(Visit == "2") %>% pull(CLS_density),
      y = data_sex %>% filter(Visit == "3") %>% pull(CLS_density),
      paired = F
    )$p.value
  )
  corrected_pvals <- p.adjust(unlist(pairwise_tests), method = "holm")
  return(corrected_pvals)
}

wilcox_pvals_F <- paired_wilcox(CLS, "F")
wilcox_pvals_M <- paired_wilcox(CLS, "M")

plot <- ggplot(CLS, aes(x = interaction(Visit, Sex), y = CLS_density, fill = interaction(Visit, Sex))) +
  geom_boxplot(outlier.shape = NA) +  # Remove outliers
  geom_jitter(color = "grey", size = 1, width = 0.2, alpha = 0.5) +  # Add individual points
  labs(title = "CLS Density Across Visits and Genders", x = "Visit and Sex", y = "# CLS per mm3") +
  theme_bw() +
  theme(plot.title = element_text(size = 10)) +
  geom_signif(
    comparisons = list(
      c("1.F", "2.F"), 
      c("1.F", "3.F"), 
      c("2.F", "3.F"),
      c("1.M", "2.M"), 
      c("1.M", "3.M"), 
      c("2.M", "3.M")
    ),
    annotations = c(
      wilcox_stars_F["visit1_vs_visit2"],
      wilcox_stars_F["visit1_vs_visit3"],
      wilcox_stars_F["visit2_vs_visit3"],
      wilcox_stars_M["visit1_vs_visit2"],
      wilcox_stars_M["visit1_vs_visit3"],
      wilcox_stars_M["visit2_vs_visit3"]
    ),
    textsize = 4,
    map_signif_level = FALSE
  )

# Add Kruskal-Wallis p-values as text
plot <- plot + 
  annotate("text", x = 1, y = max(CLS$CLS_density) * 1.1, 
           label = paste("KW Female p-value:", format(kw_pval_F, digits = 3)), 
           hjust = 0, size = 3) +
  annotate("text", x = 4, y = max(CLS$CLS_density) * 1.1, 
           label = paste("KW Male p-value:", format(kw_pval_M, digits = 3)), 
           hjust = 0, size = 3) +
  scale_fill_manual(values = sample_colours)

plot
```

## Figure 1o

```{r}
size_data <- read.csv("2D_histology_averaged_data_output.csv",sep = ",", header = TRUE)
condition_levels <- c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")

# 1. Calculate per-patient averages for each visit
patient_averages <- size_data %>%
  group_by(Donor, Visit, Sex) %>%
  summarise(Vascular_Count = mean(Vascular_count_per_area, na.rm = TRUE)) %>%
  ungroup() %>% 
  mutate(sex_visit = interaction(Visit, Sex) %>% 
           factor(labels = c("Female Visit 1", "Female Visit 2", "Female Visit 3", "Male Visit 1", "Male Visit 2", "Male Visit 3")))

# 3. Create the boxplot
plot <- ggplot(patient_averages, aes(x = sex_visit, y = Vascular_Count, fill = sex_visit)) +
  geom_boxplot(position = position_dodge(width = 0.75), outlier.shape = NA) +
  geom_jitter(size = 1, width = 0.2, alpha = 0.3) +  # Add individual points
  theme_bw() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"), 
    text = element_text(family = "serif", size = 12),
    axis.text.x = element_blank(),
    line = element_blank()
  ) +
  scale_fill_manual(values = sample_colours) +
  labs(x = "", fill = "", y = "Blood vessels per m2")

plot
```

@@

```{r f2load}
con <- qread("con_vascular.qs", nthreads = 10)
anno.vascular <- qread("anno_vascular.qs")

# For 2d,e
cm.bulk <- read.delim("Bulk_norm_counts.txt", header = T) %>% 
  .[match(unique(.$hgnc_symbol), .$hgnc_symbol), ] %>% 
  `rownames<-`(.$hgnc_symbol) %>% 
  select(-hgnc_symbol, -ensembl_gene_id, -entrezgene_id) %>% 
  scale()

colnames(cm.bulk) %<>% 
  gsub("D", "Donor_", .) %>% 
  gsub("V", "vis", .)

# Metadata
metadata.all <- read.delim("Meta_Data_WL_Select.txt", h=T, dec = ",") %>% 
  mutate(sex = factor(sex, labels = c("Female", "Male")),
         donor = gsub("D", "Donor_", recordid) %>% 
           gsub("V", "vis", .))
```

## Figure 2a

@@

## Figure 2b

@@

## Figure 2c

@@

## Figures 2d,e

```{r}
# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

anno.comb <- anno.major[!names(anno.major) %in% names(anno.vascular)] %>% 
  {factor(c(., anno.vascular))}
```

Here, we're showing the preparations to create models for prediction. This takes some time and is not run again.

```{r, eval = F}
# Get markers
markers.vascular.all <- con.major$getDifferentialGenes(groups = anno.comb, z.threshold = 1, upregulated.only = T, verbose = T)

# vEC
# Get proportions
proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(vEC, rownames(.))

# Extract markers. Please note, the marker metrics may change a little in each iteration
markers.all <- markers.vascular.all$vEC %>% 
  filter(Specificity > 0.9,
         AUC > 0.65) %>% 
  arrange(desc(AUC)) %>% 
  pull(Gene)

# Subset bulk data
cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% markers.all) %>% 
  as.matrix() %>% 
  t()

# Now, train a final model on all data for visualization and gene selection
cm.test <- cm.bulk.mat %>% 
  .[rownames(.) %in% names(proportions), ]

lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
best_lambda <- lasso_cv$lambda.min
lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)

# Predict on all samples using the final model
lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)

# Save model
list(lasso_cv = lasso_cv,
     lasso_model = lasso_model) %>%
  qsave("vEC_model.qs")

# cEC
## Get proportions
proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(cEC, rownames(.))

## Markers
markers.all <- markers.vacsular.all$cEC %>% 
  filter(Specificity > 0.9,
         AUC > 0.6) %>% 
  arrange(desc(AUC)) %>% 
  pull(Gene)

## Subset bulk data
cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% markers.all) %>% 
  as.matrix() %>% 
  t()

# Now, train a final model on all data for visualization and gene selection
cm.test <- cm.bulk.mat %>% 
  .[rownames(.) %in% names(proportions), ]

lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 40)
best_lambda <- lasso_cv$lambda.min
lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)

# Predict on all samples using the final model
lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)

# Save model
list(lasso_cv = lasso_cv,
     lasso_model = lasso_model) %>%
  qsave("cEC_model.qs")
```

### Figure 2d

```{r}
# Load and prepare
tmp <- qread("cEC_model.qs")

lasso_model <- tmp$lasso_model
best_lambda <- tmp$lasso_cv$lambda.min

cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
  as.matrix() %>% 
  t()

proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(cEC, rownames(.))

## Predict proportions
proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)

## Prepare plot
plot.df <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.trained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>%
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.untrained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(!donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

plot.df %>% 
  ggplot() +
  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
  theme_bw() +
  labs(y = "Predicted proportions (%)", title = "Predicted vECs proportions", x = "", col = "Median trends", fill = "") +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  theme(line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank())
```

Statistics

```{r}
plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::wilcox_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::wilcox_test(s1 ~ visit)
```

### Figure 2e

```{r}
# Load and prepare
tmp <- qread("vEC_model.qs")

lasso_model <- tmp$lasso_model
best_lambda <- tmp$lasso_cv$lambda.min

cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
  as.matrix() %>% 
  t()

proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(vEC, rownames(.))

## Predict proportions
proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)

plot.df <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.trained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>%
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.untrained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(!donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

plot.df %>% 
  ggplot() +
  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
  theme_bw() +
  labs(y = "Predicted proportions (%)", title = "Predicted cECs proportions", x = "", col = "Median trends", fill = "") +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  theme(line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank())
```

Statistics

```{r}
plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::wilcox_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::wilcox_test(s1 ~ visit)
```

@@

## Figure 3a

@@

## Figure 3b

@@

## Figure 3c

Here, we're plotting proportions of total cells, so we use the Conos object and the annotation for all cells first.

```{r f3c}
# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

@@

## Figures 3d,e

```{r}
# Use Cacoa to calculate proportions
cao <- Cacoa$new(data.object = con.major, 
                 cell.groups = anno.major, 
                 ref.level = "vis2", 
                 target.level = "vis3", 
                 sample.groups = con.major$samples %>% 
                   names() %>% 
                   strsplit("_") %>% 
                   sapply('[[', 3) %>% 
                   gsub(1, 2, .) %>% 
                   `names<-`(con.major$samples %>% 
                               names()),
                 sample.groups.palette = ggsci::pal_jama()(7))

anno.comb <- anno.major[!names(anno.major) %in% names(anno.immune)] %>% 
  {factor(c(., anno.immune))}
```

Here, we're showing the preparations to create models for prediction. This takes some time and is not run again.

```{r, eval = F}
# Get markers
markers.immune.all <- con.major$getDifferentialGenes(groups = anno.comb, z.threshold = 1, upregulated.only = T, verbose = T)

# vEC
# Get proportions
proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(`Early LAM`, rownames(.))

# Extract markers. Please note, the marker metrics may change a little in each iteration
markers.all <- markers.immune.all$`Early LAM` %>% 
  filter(Specificity > 0.9,
         AUC > 0.65) %>% 
  arrange(desc(AUC)) %>% 
  pull(Gene)

# Subset bulk data
cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% markers.all) %>% 
  as.matrix() %>% 
  t()

# Now, train a final model on all data for visualization and gene selection
cm.test <- cm.bulk.mat %>% 
  .[rownames(.) %in% names(proportions), ]

lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 5)
best_lambda <- lasso_cv$lambda.min
lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)

# Predict on all samples using the final model
lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)

# Save model
list(lasso_cv = lasso_cv,
     lasso_model = lasso_model) %>%
  qsave("eLAM_model.qs")

# cEC
## Get proportions
proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(LAM, rownames(.))

## Markers
markers.all <- markers.immune.all$LAM %>% 
  filter(Specificity > 0.9,
         AUC > 0.7) %>% 
  arrange(desc(AUC)) %>% 
  pull(Gene)

## Subset bulk data
cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% markers.all) %>% 
  as.matrix() %>% 
  t()

# Now, train a final model on all data for visualization and gene selection
cm.test <- cm.bulk.mat %>% 
  .[rownames(.) %in% names(proportions), ]

lasso_cv <- cv.glmnet(cm.test, proportions, alpha = 1, nfolds = 15)
best_lambda <- lasso_cv$lambda.min
lasso_model <- glmnet(cm.test, proportions, alpha = 1, lambda = best_lambda)

# Predict on all samples using the final model
lasso.predict.all <- predict(lasso_model, newx = cm.test, s = best_lambda)

# Save model
list(lasso_cv = lasso_cv,
     lasso_model = lasso_model) %>%
  qsave("LAM_model.qs")
```

### Figure 3d

```{r}
# Load and prepare
tmp <- qread("eLAM_model.qs")

lasso_model <- tmp$lasso_model
best_lambda <- tmp$lasso_cv$lambda.min

cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
  as.matrix() %>% 
  t()

proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(`Early LAM`, rownames(.))

## Predict proportions
proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)

## Prepare plot
plot.df <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.trained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>%
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.untrained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(!donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

plot.df %>% 
  ggplot() +
  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
  theme_bw() +
  labs(y = "Predicted proportions (%)", title = "Predicted eLAM proportions", x = "", col = "Median trends", fill = "") +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  theme(line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank())
```

Statistics

```{r}
plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::wilcox_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::wilcox_test(s1 ~ visit)
```

### Figure 3e

```{r}
# Load and prepare
tmp <- qread("LAM_model.qs")

lasso_model <- tmp$lasso_model
best_lambda <- tmp$lasso_cv$lambda.min

cm.bulk.mat <- cm.bulk %>% 
  as.data.frame() %>% 
  filter(rownames(.) %in% rownames(coef(lasso_model))) %>% 
  as.matrix() %>% 
  t()

proportions <- cao$.__enclos_env__$private$extractCodaData(cell.groups = anno.comb, ret.groups = F) %>% 
  {100 * ./rowSums(.)} %>% 
  as.data.frame() %$% 
  setNames(LAM, rownames(.))

## Predict proportions
proportions.bulk <-  predict(lasso_model, newx = cm.bulk.mat, s = best_lambda)

plot.df <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.trained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>%
  mutate(sex_visit = paste(sex, visit, sep = "_"))

trend.untrained <- proportions.bulk %>% 
  as.data.frame() %>%
  mutate(donor = rownames(.)) %>% 
  filter(!donor %in% names(proportions)) %>%
  mutate(visit = strsplit(donor, "_") %>% sget(3),
         sex = metadata.all$sex[match(donor, metadata.all$donor)]) %>% 
  group_by(visit, sex) %>% 
  summarize(s1 = median(s1)) %>% 
  mutate(sex_visit = paste(sex, visit, sep = "_"))

plot.df %>% 
  ggplot() +
  geom_boxplot(mapping = aes(sex_visit, s1, fill = sex_visit), outliers = F) +
  geom_jitter(mapping = aes(sex_visit, s1, col = sex_visit), alpha = 0.3, col = "black", width = 0.3) +
  theme_bw() +
  labs(y = "Predicted proportions (%)", title = "Predicted LAM  proportions", x = "", col = "Median trends", fill = "") +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Trained data"), data = trend.trained, linewidth = 1) +
  geom_line(mapping = aes(x = sex_visit, y = s1, group = sex, col = "Untrained data"), data = trend.untrained, linewidth = 1) +
  scale_color_manual(values = c("Trained data" = "black", "Untrained data" = "grey40")) +
  scale_fill_manual(values = c(colorRampPalette(c("pink","darkred"))(3), colorRampPalette(c("lightblue","blue3"))(3))) +
  theme(line = element_blank(),
        axis.text = element_text(color = "black"),
        axis.text.x = element_blank())
```

Statistics

```{r}
plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Female") %>% 
  rstatix::wilcox_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::kruskal_test(s1 ~ visit)

plot.df %>% 
  filter(sex == "Male") %>% 
  rstatix::wilcox_test(s1 ~ visit)
```

## Figure 3f

@@

# Initially, we used this approach to select genes. Since the clusters are not exactly reproducible, here we load the gene lists

# genes <- de[c("ATM","Early LAM","LAM","mono/mac")] %>%
#   {`names<-`(lapply(names(.), \(nn) mutate(.[[nn]], ct = nn)), names(.))} %>% 
#   lapply(filter, padj <= 0.05, abs(log2FoldChange) > 1) %>%
#   lapply(pull, Gene) %>% 
#   {sapply(names(.), \(nn) .[[nn]] %>% `names<-`(., rep(nn, length(.))))} %>% 
#   Reduce(c, .) %>% 
#   .[match(unique(.), .)]

genes <- read.delim("Macrophages_DEG_heatmap_genes.tsv") %$% 
  setNames(gene, cluster) %>%
  .[order(names(.))]

@@

ha <- ComplexHeatmap::Heatmap(x, 
                              name='Expression', 
                              col=pal, 
                              cluster_columns=FALSE, 
                              show_row_names=F, 
                              show_column_names=FALSE, 
                              top_annotation=tannot,
                              left_annotation=NULL,
                              border=TRUE,
                              show_column_dend = FALSE, 
                              show_row_dend = FALSE,
                              cluster_rows = F,
                              # row_km = 5, # Only used initially
                              # row_km_repeats = 500, # Only used initially
                              row_split = names(genes),
                              column_split = groups[colnames(x)] %>% strsplit("!!") %>% sapply('[[', 2) %>% unname() %>% factor(levels = c("ATM", "mono/mac", "Early LAM", "LAM")))

@@

Save order for later

```{r, eval = F}
cls %>% 
  lapply(\(y) rownames(x)[y]) %>% 
  lapply(data.frame) %>% 
  lapply(setNames, "gene") %>% 
  bind_rows(.id = "cluster") %>%
  select(gene, cluster) %>% 
  write.table("Macrophages_DEG_heatmap_genes.tsv", sep = "\t")
```

## Figures 3g,h

@@

### Figure 3g

@@

### Figure 3h

```{r}
anno <- anno.myeloid %>% 
  .[. %in% cts] %>% 
  factor()

cm.clean <- cm.merged %>% 
  .[rownames(.) %in% names(anno), ]

spc <- con$getDatasetPerCell()

lam.scores <- list(LAM.activity = genes)
seu_obj <- CreateSeuratObject(counts = cm.clean %>% Matrix::t() %>% drop0(),
                              min.cells = 0, 
                              min.features = 0)
m.scores <- UCell::AddModuleScore_UCell(seu_obj, features = lam.scores, ncores = 32)
```
